The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-05-06 paper.qmd
life is a nightmare and i hate pixel maps

# load Libraries
library(strayr)
library(tidyverse)
library(biscale)
library(RColorBrewer)
library(scales)
library(sf)
library(ggrepel)

@@

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-maps
#| fig-cap: "A visualisation of a typical choropleth map, as well as three other maps that are used to display uncertainty on the choropleth map. There is a high variance and low variance example for each type of map to show how well the technique supresses signal. Each map was created using the same data with the same base palette. At first glance, the high uncertainty bivariate and VSUP maps just look like maps with a low saturation colour palette, rather than map with high uncertainty. This visualisation makes it clear that supression methods that plot uncertainty to a second axis, such as hue, make uncertainty appear as a second variable, rather than a signal supression on our estimate. "
#| fig-subcap: 
#|   - "Choropleth Map: Low Variance"
#|   - "Choropleth Map: High Variance"
#|   - "Bivariate Map: Low Variance"
#|   - "Bivariate Map: High Variance"
#|   - "VSUP Map: Low Variance"
#|   - "VSUP Map: High Variance"
#|   - "Pixel Map: Low Variance"
#|   - "Pixel Map: High Variance"
#| layout-ncol: 2

@@

# Get map data
#sa4 <- read_absmap("sa42021")
state <- read_absmap("state2021") |>
  na.omit() |>
  dplyr::mutate(temp = 31 - 0.001*(cent_long-110)^(2) - 0.003*(cent_lat-12.52294)^(2), # trend
         notrend = rnorm(1), # no trend
         highvar = runif(9, min=2, max=4), # high variance
         lowvar = runif(9, min=0, max=2), # low variance
         ) |>
  pivot_longer(cols=highvar:lowvar, names_to = "variance_class", values_to = "variance") |>
  # add bivariate classes to data
  mutate(bitemp = cut(temp, breaks=20:28, labels=seq(8)),
         bivar = cut(variance, breaks=0:4, labels=seq(4)),
         biclass = paste(bitemp, bivar, sep="-")
  )

#sa4 <- sa4 |>
#  dplyr::mutate(temp = 30 - 0.001*(cent_long-123.7799)^(2) - 0.003*(cent_lat-12.52294)^(2), # trend
#         notrend = rnorm(1), # no trend
#         highvar = runif(1, min=4, max=5), # high variance
#         lowvar = runif(1, min=0, max=1), # low variance
#         )

# Choropleth Map
p1a <- state |>
  filter(variance_class=="lowvar") %>%
  ggplot() +
  geom_sf(aes(fill = temp, geometry = geometry)) + 
  scale_fill_distiller(type = "seq", 
                       palette = "YlOrRd", 
                       direction = 1)+
  theme_minimal() + 
  theme(legend.position = "none")

p1b <- p1a %+% filter(state, variance_class=="highvar")
# Bivariate Map
# Make bivariate palette
# Function to devalue by a certain amount
colsupress <- function(basecols, hue=1, sat=1, val=1) {
    X <- diag(c(hue, sat, val)) %*% rgb2hsv(col2rgb(basecols))
    hsv(pmin(X[1,], 1), pmin(X[2,], 1), pmin(X[3,], 1))
}

# set Base colours
basecols <- brewer.pal(8, name = "YlOrRd")
# recurvisely decrease value
v_val = 0.5
bivariatepal <- c(basecols,
                  colsupress(basecols, sat=v_val),
                  colsupress(colsupress(basecols, sat=v_val), sat=v_val),
                  colsupress(colsupress(colsupress(basecols, sat=v_val), sat=v_val), sat=v_val))
# establish levels of palette
names(bivariatepal) <- paste(rep(1:8, 4), "-" , rep(1:4, each=8), sep="")
# preview palette
show_col(bivariatepal, ncol=8)

# Bivariate maps
p2a <- state |>
  filter(variance_class=="lowvar") %>%
  ggplot() +
  geom_sf(aes(fill = biclass, geometry = geometry)) + 
  scale_fill_manual(values = bivariatepal) +
  theme_minimal() + 
  theme(legend.position = "none")

p2b <- p2a %+% filter(state, variance_class=="highvar")

@@

names(VSUP) <- paste(rep(1:8, 4), "-" , rep(1:4, each=8), sep="")
show_col(VSUP, ncol=8)

# VSUP maps
p3a <- state |>
  filter(variance_class=="lowvar") %>%
  ggplot() +
  geom_sf(aes(fill = biclass, geometry = geometry)) + 
  scale_fill_manual(values = VSUP) +
  theme_minimal() + 
  theme(legend.position = "none")

p3b <- p3a %+% filter(state, variance_class=="highvar")

# Maps I haven't been able to get working yet

# Pixel Map
library(Vizumap)
# Low variance map
state_a <- state %>%
  filter(variance_class == "lowvar") %>%
  mutate(my_id = seq(9),
         error = variance)

pixel_1a <- state_a %>%
  as.data.frame() %>%
  select(my_id, temp, error) %>%
  read.uv(estimate="temp", error="error")

pixel_2a <- state_a %>% as("Spatial")

@@

pmap_a <- build_pmap(data = pixel_1a, distribution = "normal", pixelGeo = pix_a, id = "my_id", border = pixel_2a)

p4a <- view(pmap_a) +
    scale_fill_distiller(
      type = "seq", 
      palette = "YlOrRd", 
      direction = 1) +
  scale_colour_distiller(
      type = "seq", 
      palette = "YlOrRd", 
      direction = 1) +
  theme(legend.position="none")

# High variance
state_b <- state %>%
  filter(variance_class == "highvar") %>%
  mutate(my_id = seq(9),
         error = variance)

pixel_1b <- state_b %>%
  as.data.frame() %>%
  select(my_id, temp, error) %>%
  read.uv(estimate="temp", error="error")

pixel_2b <- state_b %>% 
  as("Spatial")

@@

pmap_b <- build_pmap(data = pixel_1b, distribution = "normal", pixelGeo = pix_b, id = "my_id", border = pixel_2b)

p4b <- view(pmap_b) +
    scale_fill_distiller(
      type = "seq", 
      palette = "YlOrRd", 
      direction = 1) +
  scale_colour_distiller(
      type = "seq", 
      palette = "YlOrRd", 
      direction = 1) +
  theme(legend.position="none")

p1a
p1b
p2a
p2b
p3a
p3b
p4a
p4b
```

>>>>

# load Libraries
library(tidyverse)
library(biscale)
library(RColorBrewer)
library(scales)
library(sf)
library(ggrepel)
library(urbnmapr)

@@

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-maps
#| fig-cap: "A visualisation of a typical choropleth map, as well as three other maps that are used to display uncertainty on the choropleth map. There is a high variance and low variance example for each type of map to show how well the technique supresses signal. Each map was created using the same data with the same base palette. At first glance, the high uncertainty bivariate and VSUP maps just look like maps with a low saturation colour palette, rather than map with high uncertainty. This visualisation makes it clear that supression methods that plot uncertainty to a second axis, such as hue, make uncertainty appear as a second variable, rather than a signal supression on our estimate. "
#| fig-subcap: 
#|   - "Choropleth Map: Low Variance"
#|   - "Choropleth Map: High Variance"
#|   - "Choropleth Palette"
#|   - "Bivariate Map: Low Variance"
#|   - "Bivariate Map: High Variance"
#|   - "Bivariate Palette"
#|   - "VSUP Map: Low Variance"
#|   - "VSUP Map: High Variance"
#|   - "VSU Palette"
#|   - "Pixel Map: Low Variance"
#|   - "Pixel Map: High Variance"
#|   - "Pixel Palette"
#| layout-ncol: 3

@@

# Get map data
my_map_data <- get_urbn_map("counties", sf = TRUE) |>
  dplyr::filter(state_name=="Iowa")
centroids <- as_tibble(rgeos::gCentroid(as(my_map_data$geometry, "Spatial"), byid = TRUE))
my_map_data$cent_long <- centroids$x
my_map_data$cent_lat <- centroids$y
n <- dim(my_map_data)[1]

# Make standard Palette
# set Base colours
basecols <- brewer.pal(8, name = "YlOrRd")
#breaks <- seq(0,41, length.out = 9)
breaks <- 21:29 #c(0,15,24,25,26,27,30,35,41)
names(basecols) <- seq(8)
limits <- c(10,41)

my_map_data <- my_map_data |>
  dplyr::mutate(temp = 28 - ((scale(cent_long))^(2) + (scale(cent_lat))^(2))[,1], # trend
         notrend = rnorm(n), # no trend
         highvar = runif(n, min=2, max=4), # high variance
         lowvar = runif(n, min=0, max=2), # low variance
         ) |>
  pivot_longer(cols=highvar:lowvar, names_to = "variance_class", values_to = "variance") |>
  # add bivariate classes to data
  mutate(bitemp = cut(temp, breaks=breaks, labels=seq(8)),
         bivar = cut(variance, breaks=0:4, labels=seq(4)),
         biclass = paste(bitemp, bivar, sep="-")
  )

# Choropleth Map
p1a <- my_map_data |>
  filter(variance_class=="lowvar") %>%
  ggplot() +
  geom_sf(aes(fill = bitemp, 
              geometry = geometry)) + 
  scale_fill_manual(values = basecols) +
  #scale_fill_gradientn(colours = basecols, 
  #                     values=breaks/limits[2],
  #                     limits=limits) +
  theme_void() + 
  theme(legend.position = "none")

p1b <- p1a %+% filter(my_map_data, variance_class=="highvar")
# Bivariate Map
# Make bivariate palette
# Function to devalue by a certain amount
colsupress <- function(basecols, hue=1, sat=1, val=1) {
    X <- diag(c(hue, sat, val)) %*% rgb2hsv(col2rgb(basecols))
    hsv(pmin(X[1,], 1), pmin(X[2,], 1), pmin(X[3,], 1))
}

# recurvisely decrease value
v_val = 0.5
bivariatepal <- c(basecols,
                  colsupress(basecols, sat=v_val),
                  colsupress(colsupress(basecols, sat=v_val), sat=v_val),
                  colsupress(colsupress(colsupress(basecols, sat=v_val), sat=v_val), sat=v_val))
# establish levels of palette
names(bivariatepal) <- paste(rep(1:8, 4), "-" , rep(1:4, each=8), sep="")

# Bivariate maps
p2a <- my_map_data |>
  filter(variance_class=="lowvar") %>%
  ggplot() +
  geom_sf(aes(fill = biclass, geometry = geometry)) + 
  scale_fill_manual(values = bivariatepal) +
  theme_void() + 
  theme(legend.position = "none")

p2b <- p2a %+% filter(my_map_data, variance_class=="highvar")

@@

names(VSUP) <- paste(rep(1:8, 4), "-" , rep(1:4, each=8), sep="")

# VSUP maps
p3a <- my_map_data |>
  filter(variance_class=="lowvar") %>%
  ggplot() +
  geom_sf(aes(fill = biclass, geometry = geometry)) + 
  scale_fill_manual(values = VSUP) +
  theme_void() + 
  theme(legend.position = "none")

p3b <- p3a %+% filter(my_map_data, variance_class=="highvar")

# Pixel Map
library(Vizumap)
# Low variance map
my_map_data_a <- my_map_data %>%
  filter(variance_class == "lowvar") %>%
  mutate(my_id = seq(n),
         error = variance)

# quantile
q_a <- my_map_data_a |> 
  as.tibble() |>
  mutate(bitemp=as.numeric(bitemp)) |>
  with(data.frame(p0.05 = qnorm(0.05, mean=bitemp, sd=sqrt(variance)),
                  p0.25 = qnorm(0.25, mean=bitemp, sd=sqrt(variance)), 
                  p0.5 = qnorm(0.5, mean=bitemp, sd=sqrt(variance)), 
                  p0.75 = qnorm(0.75, mean=bitemp, sd=sqrt(variance)), 
                  p0.95 = qnorm(0.95, mean=bitemp, sd=sqrt(variance))))|>
  apply(2, function(x) ifelse(x>8, 8, x)) |>
  apply(2, function(x) ifelse(x<0, 0, x))

pixel_1a <- my_map_data_a %>%
  as.data.frame() %>%
  select(my_id, bitemp, error) %>%
  read.uv(estimate="bitemp", error="error")

pixel_2a <- my_map_data_a %>% as("Spatial")

@@

pmap_a <- build_pmap(data = pixel_1a, distribution = "discrete", pixelGeo = pix_a, id = "my_id", border = pixel_2a, q=q_a)

p4a <- view(pmap_a) +
  scale_fill_gradientn(colours = basecols) +
  scale_colour_gradientn(colours = basecols) +
  theme(legend.position="none")

# High variance
my_map_data_b <- my_map_data %>%
  filter(variance_class == "highvar") %>%
  mutate(my_id = seq(n),
         error = variance)

q_b <- my_map_data_b |> 
  as.tibble() |>
  mutate(bitemp=as.numeric(bitemp)) |>
  with(data.frame(p0.05 = qnorm(0.05, mean=bitemp, sd=sqrt(variance)),
                  p0.25 = qnorm(0.25, mean=bitemp, sd=sqrt(variance)), 
                  p0.5 = qnorm(0.5, mean=bitemp, sd=sqrt(variance)), 
                  p0.75 = qnorm(0.75, mean=bitemp, sd=sqrt(variance)), 
                  p0.95 = qnorm(0.95, mean=bitemp, sd=sqrt(variance))))|>
  apply(2, function(x) ifelse(x>8, 8, x)) |> #this will mess with the variance
  apply(2, function(x) ifelse(x<0, 0, x))

pixel_1b <- my_map_data_b %>%
  as.data.frame() %>%
  select(my_id, bitemp, error) %>%
  read.uv(estimate="bitemp", error="error")

pixel_2b <- my_map_data_b %>% 
  as("Spatial")

@@

pmap_b <- build_pmap(data = pixel_1b, distribution = "discrete", pixelGeo = pix_b, id = "my_id", border = pixel_2b, q=q_b)

p4b <- view(pmap_b) +
  scale_fill_gradientn(colours = basecols) +
  scale_colour_gradientn(colours = basecols) +
  theme(legend.position="none")

# Display plots and palettes
p1a
p1b
show_col(basecols, ncol=8, labels=FALSE)
p2a
p2b
show_col(bivariatepal, ncol=8, labels=FALSE)
p3a
p3b
show_col(VSUP, ncol=8, labels=FALSE)
p4a
p4b
show_col(basecols, ncol=8, labels=FALSE)
```
